---
title: "Visualizing time series"
output:
  html_document:
    toc: true
    theme: united
---

```{r, include=FALSE}
library(knitr)
opts_chunk$set(fig.path='figures_rmd/lec04_', fig.align='center')
```

# Import libraries and define options


```{r, message=FALSE}
library(dplyr)
library(reshape2)
library(chron)
library(ggplot2)
```

```{r}
source("GRB001.R")
```

Define options

```{r, results="hide"}
Sys.setlocale("LC_TIME","C")
options(stringsAsFactors=FALSE)
options(chron.year.abb=FALSE)
theme_set(theme_bw()) # just my preference for plots
```

# Load data


Based on past work, we can define a function that reads in the data and additionally provides several time variables.

```{r}
ReadTSeries <- function(filename, timecolumn="datetime", timeformat="%d.%m.%Y %H:%M") {
  ## read the table
  data <- read.table(filename, skip=5, header=TRUE, sep=";", check.names=FALSE)
  names(data) <- sub("[ ].*$","",names(data))
  names(data) <- sub("Date/time", timecolumn, names(data), fixed=TRUE)
  data[,timecolumn] <- as.chron(data[,timecolumn], timeformat)
  ## additional variables
  data[,"year"] <- years(data[,timecolumn])
  data[,"month"] <- months(data[,timecolumn])
  data[,"day"] <- days(data[,timecolumn])
  data[,"hour"] <- hours(data[,timecolumn])
  data[,"dayofwk"] <- weekdays(data[,timecolumn])
  data[,"daytype"] <- ifelse(data[,"dayofwk"] %in% c("Sat","Sun"),"Weekend","Weekday")
  ##
  seasons <- c("DJF","MAM","JJA","SON")
  ix.seasons <- findInterval(unclass(data[,"month"]) %% 12, seq(0,12,3))
  data[,"season"] <- factor(seasons[ix.seasons],seasons)
  ## avoid package conflict
  data[,timecolumn] <- c(unclass(data[,timecolumn]))
  ## return value
  data
}
```

Read and merge Lausanne (LAU) and Z??rich (ZUE) data:
```{r}
datapath <- "data/2013"

df <- full_join(cbind(site="LAU", ReadTSeries(file.path(datapath, "LAU.csv"))),
                cbind(site="ZUE", ReadTSeries(file.path(datapath, "ZUE.csv"))))
```

We can see that this data frame contains data from both sites.
```{r}
head(df)
tail(df)
```

Let us save this data frame for later
```{r}
saveRDS(df, "data/2013/lau-zue.rds")
```

Elongate the data frame, as before.
```{r}
lf <- melt(df, id.vars=c("site", "datetime", "season", "year","month","day","hour","dayofwk","daytype"))
```

# View variability in pollutant concentrations

Plotting your data is very good practice. Check for general trends and extreme values.

View all the measurements:
```{r, warning=FALSE, fig.width=8, fig.height=10}
ggp <- ggplot(lf)+
  facet_grid(variable~site, scale="free_y")+
  geom_line(aes(datetime, value, color=site))+
  scale_x_chron()
print(ggp)
```

In the following figures, we will summarize the measurements using non-parametric (order) statistics, which we will cover in a subsequent lecture.

## Seasonal variations

Summary by month:
```{r, warning=FALSE, fig.width=8, fig.height=10}
ggp <- ggplot(lf) +
  facet_grid(variable ~ site, scale = "free_y") +  
  geom_boxplot(aes(month, value), outlier.size = 0.5, outlier.shape = 3)
print(ggp)
```

By day type and season:
```{r, warning=FALSE, fig.width=8, fig.height=8}
ggp <- ggplot(lf %>% filter(site=="LAU" & !is.na(value))) +
  facet_grid(variable ~ season, scale = "free_y") +  
  geom_boxplot(aes(daytype, value), outlier.size = 0.5, outlier.shape = 3)
print(ggp)
```

## Diurnal variations

The following function returns a function to be used for calculation of error bars.
```{r}
Percentile <- function(perc)
  function(x) quantile(x, perc*1e-2, na.rm=TRUE)
```

Diurnal (hourly) variations in pollutant concentrations at Lausanne site:
```{r, warning=FALSE, fig.width=8, fig.height=10}
ggp <- ggplot(data=lf %>% filter(site=="LAU" & !is.na(value)),
              mapping=aes(x=hour, y=value, group=daytype, color=daytype)) +
  facet_grid(variable ~ season, scale = "free_y", drop=TRUE) +  
  geom_line(stat="summary", fun.y="median")+
  geom_errorbar(stat="summary",
                fun.ymin=Percentile(25),
                fun.ymax=Percentile(75))+
  ggtitle("LAU")
print(ggp)
```

Diurnal variations in O_3 concentrations:
```{r, warning=FALSE, fig.width=8, fig.height=4}
ggp <- ggplot(data=lf %>% filter(variable=="O3"),
              mapping=aes(x=hour, y=value, group=daytype, color=daytype)) +
  facet_grid(site ~ season, drop=TRUE) +  
  geom_line(stat="summary", fun.y="median")+
  geom_errorbar(stat="summary",
                fun.ymin=Percentile(25),
                fun.ymax=Percentile(75))+
  ggtitle("O3")
print(ggp)
```
Note that for concentrations of the same pollutant, we fix the y-scale to be the same for both rows.

Diurnal variations in NO_2 concentrations:
```{r, warning=FALSE, fig.width=8, fig.height=5}
ggp <- ggplot(data=lf %>% filter(variable=="NO2"),
              mapping=aes(x=hour, y=value, group=site, color=site)) +
  facet_grid(season ~ dayofwk, drop=TRUE) +  
  geom_line(stat="summary", fun.y="median")+
  geom_errorbar(stat="summary",
                fun.ymin=Percentile(25),
                fun.ymax=Percentile(75))+
  ggtitle("NO2")
print(ggp)
```

Why are concentrations in Lausanne higher? (hint: check location of monitoring equipment)

