---
title: "Visualizing time series"
output:
  html_document:
    toc: true
    theme: united
---

```{r, include=FALSE}
library(knitr)
opts_chunk$set(fig.path='figures_rmd/lec04_', fig.align='center')
```

# Import libraries and define options


```{r, message=FALSE}
library(dplyr)
library(reshape2)
library(chron)
library(ggplot2)
```

```{r}
source("GRB001.R")
```

Define options

```{r, results="hide"}
Sys.setlocale("LC_TIME","C")
options(stringsAsFactors=FALSE)
options(chron.year.abb=FALSE)
theme_set(theme_bw()) # just my preference for plots
```

# Load data


Based on past work, we can define a function that reads in the data and additionally provides several time variables.

R provides many functions for extraction of time information, but for atmospheric applications we often classify time periods according to season (which is not provided). We will define our own function to convert month to season:

```{r}
Month2Season <- function(month) {
  seasons <- c("DJF", "MAM", "JJA", "SON")
  index <- findInterval(month %% 12, seq(0, 12, 3))
  factor(seasons[index], seasons)
}
```

Test this new function:
```{r}
Month2Season(c(1, 3, 12))
```

Next, we define the function for importing the time series:
```{r}
ReadTSeries <- function(filename, timecolumn="datetime", timeformat="%d.%m.%Y %H:%M") {
  ## read the table
  data <- read.table(filename, skip=5, header=TRUE, sep=";", check.names=FALSE)
  names(data) <- sub("[ ].*$","",names(data)) # strip units for simplification
  names(data) <- sub("Date/time", timecolumn, names(data), fixed=TRUE)
  data[,timecolumn] <- as.chron(data[,timecolumn], timeformat)
  ## additional variables
  data[,"year"] <- years(data[,timecolumn])
  data[,"month"] <- months(data[,timecolumn])
  data[,"day"] <- days(data[,timecolumn])
  data[,"hour"] <- hours(data[,timecolumn])
  data[,"dayofwk"] <- weekdays(data[,timecolumn])
  data[,"daytype"] <- ifelse(data[,"dayofwk"] %in% c("Sat","Sun"), "Weekend", "Weekday")
  data[,"season"] <- Month2Season(unclass(data[,"month"]))
  ## avoid package conflict (for version installed in GRB001)
  data[,timecolumn] <- c(unclass(data[,timecolumn]))
  ## return value
  data
}
```

Read and merge Lausanne (LAU) and Z&uuml;rich (ZUE) data:
```{r}
datapath <- "data/2013"

df <- full_join(cbind(site="LAU", ReadTSeries(file.path(datapath, "LAU.csv"))),
                cbind(site="ZUE", ReadTSeries(file.path(datapath, "ZUE.csv"))))
```

We can see that this data frame contains data from both sites.
```{r}
head(df)
tail(df)
```

Let us save this data frame for later.
```{r}
saveRDS(df, "data/2013/lau-zue.rds")
```

Elongate the data frame, as before.
```{r}
lf <- melt(df, id.vars=c("site", "datetime", "season", "year", "month", "day", "hour", "dayofwk", "daytype"))
```

# View variability in pollutant concentrations

Plotting your data is very good practice. Check for general trends and extreme values.

View all the measurements:
```{r, warning=FALSE, fig.width=8, fig.height=10}
ggp <- ggplot(lf)+
  facet_grid(variable~site, scale="free_y")+
  geom_line(aes(datetime, value, color=site))+
  scale_x_chron()+
  theme(axis.text.x=element_text(angle=30, hjust=1))
print(ggp)
```

In the following figures, we will summarize the measurements using non-parametric (order) statistics, which we will cover in a subsequent lecture.

## Seasonal variations

Summary by month:
```{r, warning=FALSE, fig.width=8, fig.height=10}
ggp <- ggplot(lf) +
  facet_grid(variable ~ site, scale = "free_y") +
  geom_boxplot(aes(month, value), outlier.size = 0.5, outlier.shape = 3)
print(ggp)
```

By day type and season:
```{r, warning=FALSE, fig.width=8, fig.height=8}
ggp <- ggplot(lf %>% filter(site=="LAU" & !is.na(value))) +
  facet_grid(variable ~ season, scale = "free_y") +
  geom_boxplot(aes(daytype, value), outlier.size = 0.5, outlier.shape = 3)
print(ggp)
```

## Diurnal variations

The following function returns a function to be used for calculation of error bars.
```{r}
Percentile <- function(perc)
  function(x) quantile(x, perc*1e-2, na.rm=TRUE)
```

Diurnal (hourly) variations in pollutant concentrations at Lausanne site:
```{r, warning=FALSE, fig.width=8, fig.height=10}
ggp <- ggplot(data=lf %>% filter(site=="LAU" & !is.na(value)),
              mapping=aes(x=hour, y=value, group=daytype, color=daytype)) +
  facet_grid(variable ~ season, scale = "free_y", drop=TRUE) +
  geom_line(stat="summary", fun.y="median")+
  geom_errorbar(stat="summary",
                fun.ymin=Percentile(25),
                fun.ymax=Percentile(75))+
  ggtitle("LAU")
print(ggp)
```

Diurnal variations in O_3 concentrations:
```{r, warning=FALSE, fig.width=8, fig.height=4}
ggp <- ggplot(data=lf %>% filter(variable=="O3"),
              mapping=aes(x=hour, y=value, group=daytype, color=daytype)) +
  facet_grid(site ~ season, drop=TRUE) +
  geom_line(stat="summary", fun.y="median")+
  geom_errorbar(stat="summary",
                fun.ymin=Percentile(25),
                fun.ymax=Percentile(75))+
  ggtitle("O3")
print(ggp)
```
Note that for concentrations of the same pollutant, we fix the y-scale to be the same for both rows.

Diurnal variations in NO_2 concentrations:
```{r, warning=FALSE, fig.width=8, fig.height=5}
ggp <- ggplot(data=lf %>% filter(variable=="NO2"),
              mapping=aes(x=hour, y=value, group=site, color=site)) +
  facet_grid(season ~ dayofwk, drop=TRUE) +
  geom_line(stat="summary", fun.y="median")+
  geom_errorbar(stat="summary",
                fun.ymin=Percentile(25),
                fun.ymax=Percentile(75))+
  ggtitle("NO2")
print(ggp)
```

Why are concentrations in Lausanne higher? (hint: check location of monitoring equipment)

